\documentclass{beamer}
\usepackage{hyperref}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\title{Introduction to Global Shifts and Northern British Columbia}
\subtitle{Global Economic Shifts: ECON 220}
\author{Muhebullah Karimzada}
\date{Winter 2025}

\begin{document}

\maketitle

\begin{frame}{Overview of the Lecture}
    \textbf{Lecture Focus:} 
    \begin{itemize}
        \item Understanding the concept of global economic shifts.
        \item Analyzing the impact of global shifts on Northern British Columbia (NBC).
        \item Discussing how NBC can leverage its resources and address local/global challenges.
    \end{itemize}
    \textbf{Key Questions:}
    \begin{itemize}
        \item What are global economic shifts, and why do they matter?
        \item How do these shifts impact the economy of Northern British Columbia?
        \item What strategies can NBC adopt to manage these shifts?
    \end{itemize}
\end{frame}

% Slide 2: Global Economic Shifts – Key Concept
\begin{frame}{Global Economic Shifts}
    \textbf{Definition:}  
    \begin{itemize}
        \item Global economic shifts are significant and sustained changes in the global economic environment that affect production, trade, labor markets, and social structures.
        \item These shifts can arise from technological advancements, policy changes, environmental factors, and demographic trends.
    \end{itemize}
    \textbf{Example:}
    \begin{itemize}
        \item The Industrial Revolution: Shift from agrarian economies to industrial economies had lasting impacts on global trade patterns and labor markets.
    \end{itemize}
\end{frame}

% Slide 3: Why Global Shifts Matter
\begin{frame}{Why Global Shifts Matter}
    \begin{itemize}
        \item Impact global trade flows and international relations.
        \item Affect national economies, industries, and employment sectors.
        \item Influence regional economies by changing demand for resources, altering labor markets, and influencing government policies.
    \end{itemize}
    \textbf{Example:}
    \begin{itemize}
        \item 2008 Financial Crisis: Disrupted global markets, leading to changes in housing prices, unemployment, and trade flows.
    \end{itemize}
\end{frame}

% Slide 4: Key Drivers of Global Economic Shifts
\begin{frame}{Drivers of Global Economic Shifts}
    \begin{itemize}
        \item \textbf{Technological Advancements:} Automation, AI, IoT, 3D printing.
        \item \textbf{Trade and Economic Integration:} Trade agreements (e.g., NAFTA, CPTPP).
        \item \textbf{Environmental and Climate Changes:} Sustainability goals, clean technology.
        \item \textbf{Demographic Changes:} Aging populations, urbanization.
        \item \textbf{Geopolitical Dynamics:} Trade wars, economic sanctions.
    \end{itemize}
    \textbf{Discussion Question:}
    \begin{itemize}
        \item Which of these drivers do you think is most influential in shaping global economic shifts today?
    \end{itemize}
\end{frame}

% Slide 5: Technological Advancements
\begin{frame}{Technological Advancements}
    \textbf{Impact on NBC:}
    \begin{itemize}
        \item Automation in industries like forestry can improve efficiency but may displace jobs.
        \item Technologies like AI and IoT may lead to new industries, such as tech and innovation hubs in NBC.
    \end{itemize}
    \textbf{Example:}
    \begin{itemize}
        \item Automation in forestry leads to greater productivity but may reduce local job opportunities. What are the social and economic implications of this?
    \end{itemize}
\end{frame}

% Slide 6: Trade and Economic Integration
\begin{frame}{Trade and Economic Integration}
    \begin{itemize}
        \item Trade agreements such as CPTPP and NAFTA expand market access.
        \item NBC benefits from its proximity to the Asia-Pacific region, especially for LNG exports.
    \end{itemize}
    \textbf{Example:}
    \begin{itemize}
        \item The trade tension between the U.S. and China affects demand for LNG from NBC. How do such geopolitical dynamics impact regional economies like NBC’s?
    \end{itemize}
\end{frame}

% Slide 7: Environmental and Climate Changes
\begin{frame}{Environmental and Climate Changes}
    \begin{itemize}
        \item Global sustainability trends are reshaping industries like forestry, mining, and energy production.
        \item Environmental regulations add costs but can also drive innovation in green technologies.
    \end{itemize}
    \textbf{Example:}
    \begin{itemize}
        \item Hydroelectric power generation in NBC as a response to global sustainability trends. What are the advantages and challenges of focusing on renewable energy in a resource-rich region?
    \end{itemize}
\end{frame}

% Slide 8: Demographic Changes
\begin{frame}{Demographic Changes}
    \begin{itemize}
        \item Aging populations in developed countries, including Canada, increase demand for healthcare services.
        \item Outmigration of youth from NBC presents challenges for workforce sustainability.
    \end{itemize}
    \textbf{Discussion Question:}
    \begin{itemize}
        \item How can NBC address its demographic challenges and retain its youth in an increasingly competitive global economy?
    \end{itemize}
\end{frame}

% Slide 9: Geopolitical Dynamics
\begin{frame}{Geopolitical Dynamics}
    \begin{itemize}
        \item Regional conflicts, trade wars, and economic sanctions disrupt global trade and influence market access.
        \item NBC’s dependence on international markets makes it vulnerable to geopolitical changes.
    \end{itemize}
    \textbf{Example:}
    \begin{itemize}
        \item How does the U.S.-China trade war impact LNG exports from NBC? What strategies can NBC adopt to mitigate such risks?
    \end{itemize}
\end{frame}

% Slide 10: Northern British Columbia's Economic Profile
\begin{frame}{Northern British Columbia’s Economic Profile}
    \begin{itemize}
        \item Rich in natural resources: timber, natural gas, minerals.
        \item Access to the Asia-Pacific trade route via the Prince Rupert Port.
        \item Challenges: Vulnerability to commodity price volatility, aging population, and outmigration.
    \end{itemize}
    \textbf{Discussion Question:}
    \begin{itemize}
        \item What are some strategies that NBC can employ to reduce its reliance on resource extraction while maintaining economic growth?
    \end{itemize}
\end{frame}

% Slide 11: Case Study - Prince Rupert Port
\begin{frame}{Case Study: Prince Rupert Port}
    \textbf{Background:}
    \begin{itemize}
        \item Canada’s fastest-growing port, connecting to Asia-Pacific markets.
        \item Strategic location for growing trade with China, Japan, and South Korea.
    \end{itemize}
    \textbf{Opportunities:}
    \begin{itemize}
        \item Expansion of port facilities increases trade volume.
        \item Economic growth through increased export capacity, especially for LNG.
    \end{itemize}
    \textbf{Discussion Question:}
    \begin{itemize}
        \item How can NBC capitalize on its port to expand exports while addressing environmental and sustainability concerns?
    \end{itemize}
\end{frame}

% Slide 12: Indigenous Perspectives
\begin{frame}{Indigenous Perspectives}
    \begin{itemize}
        \item Indigenous communities are key stakeholders in NBC’s resource development.
        \item Indigenous-led projects, like joint ventures in LNG, provide opportunities for economic empowerment.
    \end{itemize}
    \textbf{Discussion Question:}
    \begin{itemize}
        \item What role should Indigenous communities play in NBC’s economic development? How can NBC ensure that development is beneficial for these communities?
    \end{itemize}
\end{frame}

% Slide 13: Class Discussion Questions
\begin{frame}{Class Discussion Questions}
    \begin{enumerate}
        \item What are the key global trends impacting Northern British Columbia?
        \item How can NBC balance economic growth with sustainability?
        \item What role should Indigenous communities play in NBC’s economic development?
    \end{enumerate}
\end{frame}

% Slide 14: Readings
\begin{frame}{Readings}
    \begin{itemize}
        \item Stiglitz, J. E. (2002). \textit{Globalization and Its Discontents}.
        \item Bowles, P., \& Wilson, M. (2015). \textit{Resource Communities in a Globalizing Region}.
        \item Prince Rupert Port Authority (2023). \textit{Annual Reports}.
    \end{itemize}
\end{frame}

\end{document}
