\documentclass[12pt]{article}
\usepackage{amsmath, amssymb, geometry}
\geometry{margin=1in}
\setlength{\parskip}{0.5em}
\setlength{\parindent}{0pt}

\begin{document}

\title{\textbf{Solutions: Midterm Exam 2}\\
ECON 300 -- Labour Economics}
\date{}
\maketitle

\section*{Answers and Explanations}

\begin{enumerate}

\item \textbf{B} \\
A monopolist uses $MRP$, which lies below $VMP$ and is therefore less elastic.

\item \textbf{A}

\subsection*{Economic Intuition}

Economic rent is defined as the difference between the wage a worker receives and the worker’s \textbf{reservation wage} (the minimum wage required to induce them to work).

\[
\text{Economic Rent} = W - W^{R}
\]

where:
\begin{itemize}
    \item \(W\) = actual wage
    \item \(W^{R}\) = reservation wage
\end{itemize}


If workers are earning economic rents, then:

\begin{itemize}
    \item They are being paid \textbf{more than necessary} to keep them employed.
    \item A reduction in wages \textbf{will not immediately reduce employment}.
    \item Workers will continue working as long as:
    \[
    W \geq W^{R}
    \]
\end{itemize}

\subsection*{Why Answer A is Correct}

Answer A reflects the idea that:

\begin{quote}
A firm can reduce wages \textbf{without losing workers}, as long as wages remain above the reservation wage.
\end{quote}

This is the defining property of economic rents:
\begin{itemize}
    \item Firms have some flexibility to lower wages.
    \item Workers do not leave immediately because they are still better off than their next best alternative.
\end{itemize}

\subsection*{Why Answer D is Incorrect}

Answer D states that workers would leave if wages are reduced. This is incorrect because:

\begin{itemize}
    \item Workers only leave when:
    \[
    W < W^{R}
    \]
    \item A small wage reduction \textbf{within the rent margin} does not trigger exit.
\end{itemize}

Thus, Answer D ignores the key concept of a \textbf{reservation wage threshold}.

\subsection*{Conclusion}

\[
\boxed{
\text{Economic rents imply that wages can fall without reducing employment, as long as } W \geq W^{R}.
}
\]

Therefore, \textbf{Answer A is correct}.

\item \textbf{B} \\
Monopsony implies upward-sloping supply, so $MCL > W$.

\item \textbf{A} \\
Industry supply is upward sloping as higher wages attract more workers.

\item \textbf{A} \\
Payroll taxes increase labour costs and reduce employment.

\item \textbf{C} \\
Tax burden is shared depending on elasticities.

\item \textbf{D} \\
Elasticity = $\frac{\%\Delta N}{\%\Delta W}$.

\item \textbf{E} \\
Isoprofit curves are concave due to diminishing trade-offs between safety and wages.

\item \textbf{D} \\
Efficiency wages relate to productivity channels, not directly safety.

\item \textbf{B} \\
Indifference curves reflect diminishing marginal rate of substitution.

\item \textbf{A} \\
Risk-averse workers require higher compensation for risk.

\item \textbf{A} \\
Higher isoprofit curves imply lower profits.

\item \textbf{B} \\
Slope of wage-risk locus reflects compensating differential.

\item \textbf{E} \\
Adam Smith introduced compensating wage differentials.

\item \textbf{B} \\
Workers respond to incentives; statement denying this is false.

\item \textbf{D} \\
Competitive equilibrium assumes perfect information.

\item \textbf{A} \\
Risk-averse workers choose safer jobs with lower wages.

\item \textbf{E} \\
Uber drivers accept lower wages for flexibility.

\item \textbf{B} \\
Compensation is for job characteristics, not utility itself.

\item \textbf{D} \\
Education signals ability rather than directly raising productivity.

\item \textbf{B} \\
Human capital increases productivity and earnings.

\item \textbf{C} \\
Opportunity cost is foregone earnings.

\item \textbf{C} \\
Private costs/benefits accrue to individuals.

\item \textbf{D} \\
We are looking at the earning profiles in present value terms. There is no difference between old and young individual's PV of earnings.

\item \textbf{E} \\
Perfect capital markets remove current income constraints.

\item \textbf{A} \\
Optimal schooling occurs where discounted MB = MC.

\item \textbf{A} \\
Unknown variables imply imperfect information.

\item \textbf{C} \\
Earnings rise until mid-career then decline.

\item \textbf{A} \\
$PV = \frac{Y}{r}$ approximates long streams.

\item \textbf{B} \\
Monopsony outcome: $MCL = VMP$, giving $(W_1, N_1)$.

\item \textbf{C} \\
Competitive equilibrium: supply equals demand $(W_C, N_C)$.

\item \textbf{A}



In a monopsony labour market, the firm faces an upward-sloping labour supply curve. As a result:

\begin{itemize}
    \item The firm chooses employment where:
    \[
    MCL = VMP
    \]
    
    \item This determines employment \(N_M\).
    
    \item The wage is then determined from the labour supply curve at that level of employment, giving wage \(W_M\).
\end{itemize}


Vacancies arise when there is \textbf{excess demand for labour}:
\[
\text{Vacancies} = N^d - N^s
\]

In a monopsony:

\begin{itemize}
    \item The firm sets both wage and employment optimally.
    \item It hires exactly the number of workers it wants at the chosen wage.
    \item Workers are willing to supply exactly that amount of labour at \(W_M\).
\end{itemize}

Therefore:

\[
N^d = N^s \quad \text{at the monopsony equilibrium}
\]



Even though the monopsony equilibrium is \textbf{inefficient} relative to the competitive outcome:

\begin{itemize}
    \item It is still an \textbf{equilibrium} from the firm's perspective.
    \item There is \textbf{no excess demand or excess supply}.
    \item Hence, there are \textbf{no vacancies}.
\end{itemize}

\subsection*{Conclusion}

\[
\boxed{\text{The number of vacancies in a monopsony equilibrium is zero.}}
\]

Thus, the correct answer is:

\[
\boxed{\text{A}}
\]

\item \textbf{A} \\
Area $d$ represents direct costs (tuition, books).

\item \textbf{C} \\
Area $e$ represents lifetime earnings gain from high school diploma vs high school drop-out.

\item \textbf{E} \\
Areas $b + c + d$ represent total cost of education.

\end{enumerate}

\section*{Final Answer Summary}

\begin{center}
\begin{tabular}{cccccc}
1. B & 2. A & 3. B & 4. A & 5. A & 6. C \\
7. D & 8. E & 9. D & 10. B & 11. A & 12. A \\
13. B & 14. E & 15. B & 16. D & 17. A & 18. E \\
19. B & 20. D & 21. B & 22. C & 23. C & 24. D \\
25. E & 26. A & 27. A & 28. C & 29. A & 30. B \\
31. C & 32. A & 33. A & 34. C & 35. E \\
\end{tabular}
\end{center}



\section*{Solution Key: Human Capital Investment Decision}

\subsection*{Given Information}

Johny is considering a \textbf{3-year college diploma}.

\begin{itemize}
    \item Length of program: \(3\) years
    \item Annual tuition: \(\$1{,}200\)
    \item Annual books/materials: \(\$300\)
    \item Earnings without college: \(\$8{,}000\) per year
    \item Earnings with diploma: \(\$14{,}000\) per year
    \item Working life after graduation: \(12\) years
    \item Market interest rate: \(r = 6\% = 0.06\)
\end{itemize}

\hrule
\vspace{0.5em}

\subsection*{1. Total annual direct cost of attending college}

The annual \textbf{direct cost} is:

\[
\text{Direct Cost} = \text{Tuition} + \text{Books}
\]

\[
\text{Direct Cost} = 1200 + 300 = 1500
\]

\[
\boxed{\text{Annual direct cost} = \$1{,}500}
\]

\hrule
\vspace{0.5em}

\subsection*{2. Opportunity cost per year of attending college}

The \textbf{opportunity cost} is the income Johny gives up by going to school instead of working.

\[
\text{Opportunity Cost per year} = \$8{,}000
\]

\[
\boxed{\text{Opportunity cost per year} = \$8{,}000}
\]

\hrule
\vspace{0.5em}

\subsection*{3. Total cost of the investment over the 3-year period (ignoring discounting)}

The total annual cost is:

\[
\text{Annual Total Cost} = \text{Direct Cost} + \text{Opportunity Cost}
\]

\[
\text{Annual Total Cost} = 1500 + 8000 = 9500
\]

Since the program lasts 3 years:

\[
\text{Total Cost over 3 years} = 3 \times 9500 = 28{,}500
\]

\[
\boxed{\text{Total undiscounted cost over 3 years} = \$28{,}500}
\]

\hrule
\vspace{0.5em}

\subsection*{4. Annual earnings gain from completing the diploma}

The additional annual earnings from the diploma are:

\[
\Delta Y = 14{,}000 - 8{,}000 = 6{,}000
\]

\[
\boxed{\text{Annual earnings gain} = \$6{,}000}
\]

\hrule
\vspace{0.5em}

\subsection*{5. Present value of benefits}

Johny receives the additional earnings of \(\$6{,}000\) each year for \(12\) years after graduation.

So the present value of benefits, measured \textbf{at the time of graduation}, is the present value of a 12-year annuity:

\[
PV_B^{(t=3)} = 6000 \left( \frac{1-(1.06)^{-12}}{0.06} \right)
\]

First compute:

\[
(1.06)^{12} \approx 2.0122
\]

\[
(1.06)^{-12} \approx \frac{1}{2.0122} \approx 0.49697
\]

Then:

\[
\frac{1-0.49697}{0.06} = \frac{0.50303}{0.06} \approx 8.3838
\]

So:

\[
PV_B^{(t=3)} = 6000(8.3838) \approx 50{,}302.8
\]

This is the present value \textbf{at the time Johny finishes school}. Since the schooling decision is made \textbf{today}, we must discount this amount back 3 years:

\[
PV_B^{(t=0)} = \frac{50{,}302.8}{(1.06)^3}
\]

\[
(1.06)^3 = 1.191016
\]

\[
PV_B^{(t=0)} \approx \frac{50{,}302.8}{1.191016} \approx 42{,}235.9
\]

\[
\boxed{PV(\text{Benefits}) \approx \$42{,}236}
\]

\hrule
\vspace{0.5em}

\subsection*{6. Present value of costs}

Johny incurs total annual cost of \(\$9{,}500\) for 3 years. These costs occur during years 1, 2, and 3.

Thus the present value of costs is:

\[
PV_C = \frac{9500}{1.06} + \frac{9500}{(1.06)^2} + \frac{9500}{(1.06)^3}
\]

Compute each term:

\[
\frac{9500}{1.06} \approx 8962.26
\]

\[
\frac{9500}{1.1236} \approx 8454.97
\]

\[
\frac{9500}{1.191016} \approx 7974.71
\]

Add them:

\[
PV_C \approx 8962.26 + 8454.97 + 7974.71 = 25{,}391.94
\]

\[
\boxed{PV(\text{Costs}) \approx \$25{,}392}
\]

\hrule
\vspace{0.5em}

\subsection*{7. Net Present Value (NPV) decision}

The net present value is:

\[
NPV = PV(\text{Benefits}) - PV(\text{Costs})
\]

\[
NPV = 42{,}235.9 - 25{,}391.94 = 16{,}843.96
\]

\[
\boxed{NPV \approx \$16{,}844}
\]

Since:

\[
NPV > 0
\]

Johny \textbf{should invest} in the diploma.

\[
\boxed{\text{Yes, Johny should undertake the investment.}}
\]

\hrule
\vspace{0.5em}

\subsection*{8. Effect of a higher interest rate (\(r=8\%\))}

If the market interest rate rises, future earnings are discounted more heavily.

This means:

\begin{itemize}
    \item The present value of future benefits falls.
    \item The attractiveness of education decreases.
\end{itemize}

Now compute the present value of benefits at \(8\%\):

\[
PV_B^{(t=3)} = 6000 \left( \frac{1-(1.08)^{-12}}{0.08} \right)
\]

\[
(1.08)^{12} \approx 2.5182
\]

\[
(1.08)^{-12} \approx 0.3971
\]

\[
\frac{1-0.3971}{0.08} = \frac{0.6029}{0.08} \approx 7.5363
\]

\[
PV_B^{(t=3)} \approx 6000(7.5363) = 45{,}217.8
\]

Discounting back 3 years:

\[
PV_B^{(t=0)} = \frac{45{,}217.8}{(1.08)^3}
\]

\[
(1.08)^3 = 1.2597
\]

\[
PV_B^{(t=0)} \approx \frac{45{,}217.8}{1.2597} \approx 35{,}895.7
\]

Now compute the present value of costs:

\[
PV_C = \frac{9500}{1.08} + \frac{9500}{(1.08)^2} + \frac{9500}{(1.08)^3}
\]

\[
\frac{9500}{1.08} \approx 8796.30
\]

\[
\frac{9500}{1.1664} \approx 8144.15
\]

\[
\frac{9500}{1.2597} \approx 7541.48
\]

\[
PV_C \approx 8796.30 + 8144.15 + 7541.48 = 24{,}481.93
\]

Thus:

\[
NPV_{8\%} \approx 35{,}895.7 - 24{,}481.93 = 11{,}413.77
\]

So the NPV is still positive, but smaller.

\[
\boxed{\text{A higher interest rate reduces the attractiveness of education, but Johny should still invest.}}
\]

\hrule
\vspace{0.5em}

\subsection*{9. Approximate internal rate of return (IRR)}

The IRR is the discount rate that makes:

\[
PV(\text{Benefits}) = PV(\text{Costs})
\]

We already know:

\begin{itemize}
    \item At \(6\%\), NPV is positive.
    \item At \(8\%\), NPV is still positive.
\end{itemize}

So the IRR must be \textbf{above 8\%}.

Try \(12\%\):

\[
PV_B^{(t=3)} = 6000\left(\frac{1-(1.12)^{-12}}{0.12}\right)
\]

\[
(1.12)^{12} \approx 3.89598
\]

\[
(1.12)^{-12} \approx 0.2567
\]

\[
\frac{1-0.2567}{0.12} = \frac{0.7433}{0.12} \approx 6.194
\]

\[
PV_B^{(t=3)} \approx 6000(6.194)=37{,}164
\]

Discount back 3 years:

\[
PV_B^{(t=0)} = \frac{37{,}164}{(1.12)^3}
\]

\[
(1.12)^3 = 1.404928
\]

\[
PV_B^{(t=0)} \approx 26{,}453
\]

Now costs at 12\%:

\[
PV_C = \frac{9500}{1.12} + \frac{9500}{(1.12)^2} + \frac{9500}{(1.12)^3}
\]

\[
\approx 8482.14 + 7573.34 + 6761.91 = 22{,}817.39
\]

So:

\[
NPV_{12\%} \approx 26{,}453 - 22{,}817 = 3{,}636
\]

Still positive, so IRR is above \(12\%\).

Try \(15\%\):

Benefits:
\[
PV_B^{(t=3)} = 6000\left(\frac{1-(1.15)^{-12}}{0.15}\right)
\]

\[
(1.15)^{12}\approx 5.3503,\qquad (1.15)^{-12}\approx 0.1869
\]

\[
\frac{1-0.1869}{0.15} \approx 5.4207
\]

\[
PV_B^{(t=3)} \approx 6000(5.4207)=32{,}524.2
\]

Discount back 3 years:

\[
(1.15)^3=1.520875
\]

\[
PV_B^{(t=0)}\approx \frac{32{,}524.2}{1.520875}\approx 21{,}385
\]

Costs:
\[
PV_C=\frac{9500}{1.15}+\frac{9500}{(1.15)^2}+\frac{9500}{(1.15)^3}
\]

\[
\approx 8260.87+7183.37+6246.41=21{,}690.65
\]

Now NPV is slightly negative:

\[
NPV_{15\%}\approx 21{,}385-21{,}691\approx -306
\]

So the IRR is between \(14\%\) and \(15\%\), very close to \(15\%\).

\[
\boxed{IRR \approx 14.8\% \text{ to } 15\%}
\]

\hrule
\vspace{0.5em}

\subsection*{10. Comparative statics}

\subsubsection*{(a) If expected post-education earnings fall}

Then:

\[
\Delta Y = \text{earnings with diploma} - \text{earnings without diploma}
\]

becomes smaller.

Therefore:
\begin{itemize}
    \item present value of benefits falls,
    \item NPV falls,
    \item IRR falls.
\end{itemize}

So Johny becomes \textbf{less likely} to invest.

\[
\boxed{\text{Lower post-education earnings reduce the attractiveness of schooling.}}
\]

\subsubsection*{(b) If tuition fees increase}

Higher tuition raises direct cost \(D\), so:
\begin{itemize}
    \item marginal cost rises,
    \item present value of costs rises,
    \item NPV falls.
\end{itemize}

Thus Johny becomes \textbf{less likely} to invest.

\[
\boxed{\text{Higher tuition reduces the incentive to invest in education.}}
\]

\subsubsection*{(c) If his working life becomes shorter}

A shorter working life means fewer years over which to earn the higher post-education income.

Thus:
\begin{itemize}
    \item present value of benefits falls,
    \item NPV falls,
    \item IRR falls.
\end{itemize}

So Johny becomes \textbf{less likely} to invest.

\[
\boxed{\text{A shorter working life reduces the return to education.}}
\]

\hrule
\vspace{0.8em}

\subsection*{Final Conclusion}

Johny should invest in the 3-year diploma because:

\[
PV(\text{Benefits}) > PV(\text{Costs})
\]

and

\[
NPV > 0
\]

The estimated internal rate of return is approximately:

\[
\boxed{IRR \approx 15\%}
\]

which is well above the market interest rate of \(6\%\).

\[
\boxed{\text{Therefore, the diploma is a worthwhile human capital investment.}}
\]

\end{document}