\documentclass[12pt]{article}
\usepackage{amsmath, amssymb}
\usepackage{geometry}
\usepackage{enumitem}
\geometry{margin=1in}
\setlength{\parskip}{0.6em}
\setlength{\parindent}{0pt}

\begin{document}

\section*{Problem: (15 points) }

Johny is deciding whether to pursue a \textbf{three-year college diploma}. He evaluates this decision purely as an \textbf{investment in human capital}.


\begin{itemize}
    \item Length of program: \textbf{3 years}
    
    \item Annual direct costs:
    \begin{itemize}
        \item Tuition: \$1{,}200
        \item Books and materials: \$300
    \end{itemize}
    
    \item If Johny does \textbf{not} attend college: -- He can earn \$8{,}000 per year working immediately
  
    
    \item If Johny \textbf{completes} the diploma: -- He will earn \$14{,}000 per year

    
    \item Time horizon: -- Johny will work for \textbf{12 years after graduation}
    
    
    \item Market interest rate:  \( r = 6\% \)
    
\end{itemize}

\subsection*{Questions}

\begin{enumerate}
    \item Calculate Johny’s \textbf{total annual cost of attending college}.
    
    \item Compute the \textbf{opportunity cost per year} of attending college.
    
    \item Calculate the \textbf{total cost of the investment} over the 3-year period (ignore discounting).
    
    \item Calculate the \textbf{annual earnings gain} from completing the diploma.
    
    \item Compute the \textbf{present value of benefits} (additional earnings after graduation).
    
    \item Compute the \textbf{present value of costs} (including both direct and opportunity costs).
    
    \item Should Johny invest in the diploma based on the \textbf{net present value} rule?
    
    \item Suppose the market interest rate rises to \(8\%\). Explain how this affects Johny’s decision.
    
    \item Determine approximately the \textbf{internal rate of return (IRR)} of the investment.
    
    \item Explain how Johny’s decision would change if:

    \begin{enumerate}[label=(\alph*)]
        \item  His expected post-education earnings fall
        \item Tuition fees increase
        \item His working life becomes shorter
    \end{enumerate}

\end{enumerate}

\end{document}