\documentclass[11pt]{article}
\usepackage[margin=1in]{geometry}
\usepackage{amsmath, amssymb, amsthm}
\usepackage{graphicx, booktabs}
\usepackage[shortlabels]{enumitem}
\usepackage{hyperref}
\usepackage{fancyhdr}
\usepackage{microtype}

\newtheorem{theorem}{Theorem}[section]
\newtheorem{proposition}[theorem]{Proposition}
\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{example}[theorem]{Example}
\theoremstyle{remark}
\newtheorem*{remark}{Remark}

\newcommand{\MC}{\mathrm{MC}}
\newcommand{\MR}{\mathrm{MR}}
\newcommand{\MU}{\mathrm{MU}}
\newcommand{\MRS}{\mathrm{MRS}}
\newcommand{\CS}{\mathrm{CS}}
\newcommand{\PS}{\mathrm{PS}}

\pagestyle{fancy}
\fancyhf{}
\lhead{ECON 101 --- Introduction to Economics}
\chead{Lecture 1}
\rhead{Fall 2026}
\cfoot{\thepage}

\begin{document}

\begin{center}
  {\LARGE\bfseries ECON 101: Introduction to Economics}\\[0.4em]
  {\large Lecture 1: What is Economics?}
\end{center}
\medskip\hrule\medskip

\section{Motivation}

Economics studies how individuals, firms, and societies allocate scarce resources.

\section{Core Concepts}

\begin{definition}
  \textbf{Scarcity} is the condition in which wants exceed available resources.
\end{definition}

\begin{definition}
  An \textbf{opportunity cost} is the value of the next-best alternative forgone when a choice is made.
\end{definition}

\begin{example}
  Choosing to attend university full-time means forgoing the wages you could have earned working full-time.
\end{example}

\section{The Economic Way of Thinking}

Key questions every economic analysis asks:
\begin{enumerate}[(1)]
  \item What are the \textit{costs and benefits}?
  \item What happens \textit{at the margin}?
  \item What are the \textit{incentive effects}?
\end{enumerate}

\section{Microeconomics vs.\ Macroeconomics}

\begin{itemize}
  \item \textbf{Microeconomics:} individual decisions, markets, prices.
  \item \textbf{Macroeconomics:} aggregate output, unemployment, inflation, growth.
\end{itemize}

\section{Summary}

\begin{itemize}
  \item Economics is the study of choice under scarcity.
  \item Opportunity cost is the fundamental concept.
  \item We use models to reason carefully about trade-offs.
\end{itemize}

\end{document}
