\documentclass[12pt]{article}
\usepackage{amsmath, amssymb, geometry, graphicx}
\geometry{margin=1in}
\setlength{\parskip}{0.3em}
\setlength{\parindent}{0pt}

\begin{document}

\title{\textbf{Empirical Assignment}\\
ECON 300: Labour Economics}
\date{\textbf{Due: April 07,  2026}}
\maketitle

\section*{Objective}

This assignment applies human capital theory and compensating wage differentials using real data.

\section*{Dataset}

You are provided with a dataset containing:
\begin{itemize}
    \item Wage
    \item Education
    \item Experience
    \item Job risk indicator
\end{itemize}

\section*{Part I: Descriptive Analysis (10 marks)}

\begin{enumerate}
    \item Compute summary statistics.
    \item Plot wage vs education.
    \item Plot wage distribution.
    \item Briefly describe patterns.
\end{enumerate}

\section*{Part II: Human Capital Model (20 marks)}

Estimate:

\[
\ln(wage_i) = \beta_0 + \beta_1 education_i + \beta_2 experience_i + \beta_3 experience_i^2 + u_i
\]

\begin{enumerate}
    \item Interpret coefficients.
    \item Explain diminishing returns.
    \item Compute marginal effect of experience.
\end{enumerate}

\section*{Part III: Compensating Differentials (20 marks)}

Estimate:

\[
\ln(wage_i) = \beta_0 + \beta_1 education_i + \beta_2 experience_i + \beta_3 risk_i + u_i
\]

\begin{enumerate}
    \item Interpret the coefficient on risk.
    \item Does the data support compensating wage differentials?
\end{enumerate}



\section*{Part IV: Policy Discussion (15 marks)}

\begin{enumerate}
    \item Should education be subsidized?
    \item Should risky jobs be regulated?
\end{enumerate}

\section*{Submission}

Submit:
\begin{itemize}
    \item Report (max 5 pages)
    \item Graphs and tables
    \item Code/Excel Sheet 
    \item\textbf{ Late Submission penalty: $10\%$ per day.}
\end{itemize}

\end{document}