\documentclass[12pt]{article}
\usepackage{geometry}
\geometry{a4paper, margin=1in}
\usepackage{graphicx}
\usepackage{url}
\usepackage{hyperref}
\hypersetup{
    colorlinks=true,
    linkcolor=blue,
    filecolor=magenta,      
    urlcolor=cyan,
}
\usepackage{enumitem}
\usepackage{titlesec}
\titleformat{\section}{\large\bfseries}{\thesection}{1em}{}
\titleformat{\subsection}{\bfseries}{\thesubsection}{1em}{}

\title{ECON 204: Contemporary Economic Issues \\ Term Project}
%\author{Department of Economics \\ University of Northern British Columbia}
\date{Winter 2026}

\begin{document}

\maketitle

\section*{Project Overview}
In this term project, you will explore a contemporary economic issue affecting Canada. The goal is to analyze the issue from an economic perspective, using relevant theories, data, and policy frameworks. You will work in pairs to research, write a report, and present your findings to the class. This project will help you develop critical thinking, research, and communication skills while deepening your understanding of Canada's economic landscape.

\section*{Suggested Topics}
Choose one of the following topics for your project:
\begin{enumerate}[label=\arabic*.]
  %  \item \textbf{Housing Affordability Crisis in Canada}: Analyze the causes, consequences, and potential solutions to rising housing prices and rent in major Canadian cities.
    \item \textbf{Climate Change and Carbon Pricing}: Evaluate the effectiveness of Canada’s carbon pricing policies in reducing emissions and their economic impact on households and businesses.
    \item \textbf{Income Inequality in Canada}: Examine trends in income and wealth inequality, their drivers, and the role of government policies in addressing the issue.
    \item \textbf{The Future of Work in Canada}: Explore the impact of automation, artificial intelligence, and remote work on Canada’s labor market and economy.
    \item \textbf{Indigenous Economic Development}: Investigate the economic challenges and opportunities faced by Indigenous communities in Canada and the role of reconciliation in fostering economic growth.
    \item \textbf{Canada’s Aging Population}: Analyze the economic implications of an aging population, including impacts on healthcare, pensions, and labor markets.
    \item \textbf{The Gig Economy and Labor Rights}: Assess the growth of the gig economy in Canada and its implications for workers’ rights, wages, and social safety nets.
    \item \textbf{Canada’s Trade Relations in a Changing Global Economy}: Examine the impact of shifting global trade dynamics (e.g., US-China tensions, NAFTA/USMCA) on Canada’s economy.
   % \item \textbf{Inflation and Cost of Living Pressures}: Investigate the causes of recent inflation trends in Canada and their impact on households and businesses.
    \item \textbf{The Role of Immigration in Canada’s Economy}: Analyze the economic contributions of immigrants to Canada and the challenges of integrating newcomers into the labor market.
    \item \textbf{Healthcare Sustainability in Canada}: Examine the economic challenges facing Canada’s public healthcare system, including rising costs, wait times, and the impact of an aging population. Propose solutions to ensure its long-term sustainability.
\end{enumerate}

\section*{Project Requirements}
\subsection*{1. Written Report (5-8 pages excluding tables, figures, diagrams and bibliography)}
Your report should include the following sections:
\begin{itemize}
    \item \textbf{Introduction}: Clearly define the issue and its relevance to Canada’s economy.
    \item \textbf{Literature Review}: Summarize key theories, studies, and data related to the topic.
    \item \textbf{Analysis}: Use economic concepts and data to analyze the issue. Include graphs, charts, or tables where appropriate.
    \item \textbf{Policy Recommendations}: Propose evidence-based solutions or policy interventions to address the issue.
    \item \textbf{Conclusion}: Summarize your findings and their broader implications.
    \item \textbf{References}: Cite all sources using APA or MLA format.
\end{itemize}

\subsection*{2. Presentation (20-30 minutes)}
Your presentation should:
\begin{itemize}
    \item Provide a concise overview of your topic and key findings.
    \item Use visual aids (e.g., PowerPoint) to enhance your delivery.
    \item Engage the audience with clear explanations and examples.
    \item Allocate time for a Q\&A session at the end.
\end{itemize}
\newpage
\section*{Evaluation Criteria}
\textbf{Your final draft (40\%) will be graded based on the following:}
\begin{itemize}
    \item \textbf{Depth of Analysis} (30\%): Application of economic theories and concepts.
    \item \textbf{Quality of Research} (30\%): Use of credible sources, data, and evidence.
    \item \textbf{Clarity and Organization} (20\%): Logical flow and readability of the report.
    \item \textbf{Policy Recommendations} (20\%): Feasibility and relevance of proposed solutions.
\end{itemize}
\vspace{1cm}
\textbf{Your presentation (40\%) will be graded based on the following:}
\begin{itemize}
    \item \textbf{Depth of Analysis} (30\%): Application of economic theories and concepts.
    \item \textbf{Quality of Research} (30\%): Use of credible sources, data, and evidence.
    \item \textbf{Clarity and Organization} (20\%): Logical flow and readability of the presentation.

    \item \textbf{Presentation Skills} (20\%): Engagement, clarity, and professionalism during the presentation.
\end{itemize}
\vspace{0.5cm}
\textbf{The remaining 20\% will be for participation (asking questions during presentations).}
\section*{Important Dates}
\begin{itemize}
    \item \textbf{Topic Selection Deadline}: February 10, 2026

    \item \textbf{Draft Report Submission}: March 10, 2026 
    \item \textbf{Final Report Submission}: March 31, 2026
    \item \textbf{Presentation Dates}: March 24- April 9, 2026 (Exact dates to be assigned)
\end{itemize}

\section*{Tips for Success}
\begin{itemize}
    \item Start early and divide tasks equally between group members.
    \item Use reliable sources such as Statistics Canada, academic journals, and government reports.
    \item Practice your presentation to ensure it fits within the time limit and flows smoothly.

\end{itemize}

\section*{Additional Resources}
\begin{itemize}
    \item Statistics Canada: \url{https://www.statcan.gc.ca/}
    \item Bank of Canada: \url{https://www.bankofcanada.ca/}
    \item Government of Canada Publications: \url{https://www.canada.ca/en/services/finance/publications.html}
\end{itemize}

\vspace{2cm}
\centering
\textbf{Good luck!} I look forward to seeing your insights on these important economic issues. If you have any questions, feel free to reach out.

\end{document}