\documentclass{article}
\usepackage{amsmath, graphicx}
\begin{document}

\title{Aggregate Demand and Aggregate Supply}
\author{ECON 101: Macroeconomics}
\date{}
\maketitle

\section*{Exercise 1: Calculating AD and AS Equilibrium}
The aggregate demand (AD) and short-run aggregate supply (SRAS) for an economy are given by the following equations:
\begin{align*}
    AD: & \quad Y = 5000 - 200P \\
    SRAS: & \quad Y = 2500 + 100P
\end{align*}
where \(Y\) is real GDP and \(P\) is the price level.

\textbf{Questions:}
\begin{enumerate}
    \item Find the equilibrium price level (\(P^*\)) and equilibrium output (\(Y^*\)).
    \item Suppose consumer confidence decreases, shifting AD to \( Y = 4800 - 200P \). Find the new equilibrium.
    \item Illustrate the initial and new equilibrium graphically using AD and AS curves.
\end{enumerate}



\section*{Exercise 2: Graphing the Effects of a Negative Supply Shock}
Consider an economy in equilibrium at \(Y = 5500\) and \(P = 50\). \\

\begin{align*}
    AD: & \quad Y = 8000 - 50P 
\end{align*}

A supply shock (e.g., an oil price increase) shifts SRAS leftward from:
\begin{align*}
    SRAS_1: & \quad Y = 3000 + 50P \\
    SRAS_2: & \quad Y = 2800 + 50P
\end{align*}

\textbf{Questions:}
\begin{enumerate}
    \item Determine the new equilibrium price level and output.
    \item Illustrate the shift in SRAS using a graph.
    \item What happens to inflation and unemployment after this shock?
    \item If the central bank responds by increasing the money supply, how would AD shift?
\end{enumerate}

\section*{Exercise 3: Fiscal Policy and AD Shifts}
Suppose the government implements an expansionary fiscal policy by increasing spending by \$200 billion. The marginal propensity to consume (MPC) is 0.8.

\textbf{Questions:}
\begin{enumerate}
    \item Calculate the total increase in GDP using the spending multiplier formula:
    \begin{align*}
        \text{Multiplier} &= \frac{1}{1 - MPC} \\
        \Delta Y &= \text{Multiplier} \times \Delta G
    \end{align*}
    \item If the initial AD curve was \( Y = 4000 - 100P \), write the new AD equation after the fiscal policy.
    \item Illustrate the impact on the AD-AS diagram, showing the shift in AD.
    \item What happens to the unemployment rate if output increases?
\end{enumerate}

\section*{Exercise 4: Long-Run Adjustment to a Recession}
An economy is experiencing a \textbf{recessionary gap} with actual GDP at \(2700\) and full-employment GDP at \(3000\). Assume wages are \textbf{sticky in the short run} but adjust over time.

\textbf{Questions:}
\begin{enumerate}
    \item Show the initial recessionary gap on an AD-AS diagram.
    \item Over time, explain and graph how the SRAS curve adjusts to restore full employment.
    \item What role does government intervention play in accelerating this adjustment?
    \item How does this compare to classical vs. Keynesian views on policy?
\end{enumerate}

\end{document}
