\documentclass[12pt]{article}
\usepackage[a4paper,margin=1in]{geometry}
\usepackage{titlesec}
\usepackage{enumitem}
\usepackage{setspace}
\setstretch{1.15}

\titleformat{\section}{\large\bfseries}{\thesection.}{0.5em}{}
\titleformat{\subsection}{\normalsize\bfseries}{\thesubsection.}{0.5em}{}

\begin{document}

\begin{center}
    \Large\textbf{UNBC Economics Student Recruitment Video Scripts and Storyboard Guide}
\end{center}

\vspace{1em}

\section{Why I Chose Economics at UNBC}

\subsection*{Script}
\begin{quote}
``Hi! I’m [Name], and I’m studying Economics here at the University of Northern British Columbia. 
I chose Economics because I wanted to understand why things happen — why prices rise, why jobs change, and how decisions shape our communities. 
At UNBC, I get to explore these questions in small classes where professors actually know me by name. 
What’s amazing is how Economics connects math, people, and real-world issues — right here in northern B.C. 
If you’re curious about how the world really works, come study Economics at UNBC — where big ideas meet a close-knit community.''
\end{quote}

\subsection*{Storyboard}
\begin{itemize}[leftmargin=*]
    \item \textbf{Locations:} UNBC Prince George campus, classroom, library.
    \item \textbf{0:00–0:10:} Student walking on campus. Introduction.
    \item \textbf{0:10–0:30:} Classroom discussion. 
    \item \textbf{0:30–0:50:} Whiteboard, group work.
    \item \textbf{0:50–1:00:} Outdoor closing shot with UNBC logo.
\end{itemize}

\section{What Surprised Me About Studying Economics at UNBC}

\subsection*{Script}
\begin{quote}
``When I first enrolled in Economics at UNBC, I thought it was just about numbers and graphs — but it’s so much more than that. 
We talk about housing affordability in northern communities, climate change, and local labour markets. 
The professors make everything real and practical, and we use data to answer real-world questions. 
I was surprised by how much Economics is about problem-solving and thinking critically about our society. 
If you want to make sense of the world around you, you’ll love studying Economics at UNBC.''
\end{quote}

\subsection*{Storyboard}
\begin{itemize}[leftmargin=*]
    \item \textbf{Locations:} Classroom, study lounge, students collaborating.
    \item \textbf{0:00–0:10:} Student looking at laptop.
    \item \textbf{0:10–0:30:} Charts on housing and climate.
    \item \textbf{0:30–0:50:} Interaction with professor.
    \item \textbf{0:50–1:00:} Group discussion and closing line.
\end{itemize}

\section{How Economics Changed How I See the World}

\subsection*{Script}
\begin{quote}
``Before I started Economics at UNBC, I didn’t realize how much the economy shapes everything — from grocery prices to job opportunities. 
Now, when I read the news, I actually understand what’s happening and why. 
Studying Economics here taught me to think critically about policy decisions, local industries, and even my own choices. 
It’s like seeing the hidden side of how the world works. 
UNBC helped me connect theory to what’s really happening in B.C. and beyond.''
\end{quote}

\subsection*{Storyboard}
\begin{itemize}[leftmargin=*]
    \item \textbf{Locations:} News headlines, classroom, city shots.
    \item \textbf{0:00–0:10:} Student reading news.
    \item \textbf{0:10–0:30:} Data analysis scene.
    \item \textbf{0:30–0:50:} Walking near UNBC sign.
    \item \textbf{0:50–1:00:} Confident closing line.
\end{itemize}

\section{My Favourite Economics Course at UNBC}

\subsection*{Script}
\begin{quote}
``One of my favourite courses at UNBC is Labour Economics. 
We learn about wages, job markets, and how technology changes work — topics that matter in every community. 
The best part is how interactive it is — small classes mean we actually debate real policy issues and share ideas. 
Every course in Economics helps me see how data, theory, and people all fit together. 
It’s learning that feels connected to the real world — and that’s what makes UNBC special.''
\end{quote}

\subsection*{Storyboard}
\begin{itemize}[leftmargin=*]
    \item \textbf{Locations:} Classroom, whiteboard, student debate.
    \item \textbf{0:00–0:10:} Writing on board.
    \item \textbf{0:10–0:30:} Students debating.
    \item \textbf{0:30–0:50:} Professor interaction.
    \item \textbf{0:50–1:00:} Closing smile.
\end{itemize}

\section{The Economics Community at UNBC}

\subsection*{Script}
\begin{quote}
``What I love about UNBC is the sense of community — especially in Economics. 
Professors really care about us and encourage us to ask questions and explore new ideas. 
We have opportunities to work as research assistants, join discussions, and collaborate on projects that matter to our region. 
Because classes are small, you build real connections with classmates and professors. 
It’s not just a program — it’s a supportive network that helps you grow.''
\end{quote}

\subsection*{Storyboard}
\begin{itemize}[leftmargin=*]
    \item \textbf{Locations:} Study group, café, office hour.
    \item \textbf{0:00–0:10:} Group shot, laughter.
    \item \textbf{0:10–0:30:} Student with professor.
    \item \textbf{0:30–0:50:} Research project poster.
    \item \textbf{0:50–1:00:} Group waving at camera.
\end{itemize}

\section{Where Economics at UNBC Can Take You}

\subsection*{Script}
\begin{quote}
``An Economics degree from UNBC opens so many doors. 
Our grads work in government, business, finance, data analysis, and policy — both here in northern B.C. and across Canada. 
What I’ve learned at UNBC — critical thinking, problem-solving, and understanding data — are skills every employer looks for. 
Economics gives you the flexibility to work anywhere, and UNBC gives you the personal guidance to get there. 
It’s a program that prepares you for impact.''
\end{quote}

\subsection*{Storyboard}
\begin{itemize}[leftmargin=*]
    \item \textbf{Locations:} Computer lab, downtown, co-op site.
    \item \textbf{0:00–0:10:} Student analyzing data.
    \item \textbf{0:10–0:30:} Career fair interaction.
    \item \textbf{0:30–0:50:} Walking out of building.
    \item \textbf{0:50–1:00:} Closing smile.
\end{itemize}

\section{Economics and Real-World Impact at UNBC}

\subsection*{Script}
\begin{quote}
``What makes Economics at UNBC unique is how connected it is to real-world challenges. 
We look at issues like housing, environmental sustainability, and community development — right here in northern Canada. 
You see how economics can help design better policies and improve people’s lives. 
It’s not just theory — it’s about making change that matters. 
That’s why I’m proud to study Economics at the University of Northern British Columbia.''
\end{quote}

\subsection*{Storyboard}
\begin{itemize}[leftmargin=*]
    \item \textbf{Locations:} Northern landscape, classroom, forest view.
    \item \textbf{0:00–0:10:} Panoramic campus shot.
    \item \textbf{0:10–0:30:} Presentation on environment or housing.
    \item \textbf{0:30–0:50:} Students discussing solutions.
    \item \textbf{0:50–1:00:} Outdoor closing shot.
\end{itemize}

\section*{General Filming Tips}
\begin{itemize}[leftmargin=*]
    \item Use natural light and show UNBC’s scenic campus.
    \item Feature real students — authentic voices are key.
    \item Highlight diversity and inclusion.
    \item End each video with the UNBC logo and tagline: ``Canada’s Green University''.
    \item Add upbeat background music and captions.
\end{itemize}

\end{document}
