\documentclass[12pt]{article}

\usepackage[margin=1in]{geometry}
\usepackage{amsmath, amssymb}
\usepackage{setspace}
\usepackage{enumitem}
\usepackage{titlesec}
\usepackage{hyperref}

\setstretch{1.25}
\setlist[itemize]{noitemsep, topsep=2pt}
\setlist[enumerate]{noitemsep, topsep=2pt}

\titleformat{\section}{\large\bfseries}{\thesection.}{0.5em}{}
\titleformat{\subsection}{\normalsize\bfseries}{\thesubsection}{0.5em}{}

\title{\textbf{Core Research Questions in Modern Macroeconomics}}
\date{}
\author{}

\begin{document}
\maketitle

\section*{Overview}
Modern macroeconomics focuses on quantitatively important questions about economic fluctuations, growth, policy effectiveness, and distributional outcomes. The following research areas represent the core questions studied in top journals.

\section{What Drives Business Cycles and Economic Flucuations?}

\textbf{Central Question}

Why do output, employment, and investment fluctuate over time?

\textbf{Sub-questions}
\begin{itemize}
\item Are cycles driven by productivity shocks, financial shocks, or demand shocks?
\item How important are expectations and uncertainty?
\item What explains the severity of recessions?
\end{itemize}

\textbf{Methods}
\begin{itemize}
\item DSGE models
\item Structural VARs
\item Micro-founded macro models
\end{itemize}

\textbf{Why top journals care}

Understanding fluctuations is essential for stabilization policy and welfare.

\section{How Effective is Monetary Policy?}

\textbf{Central Question}

How do interest rate changes affect inflation, output, and financial stability?

\textbf{Sub-questions}
\begin{itemize}
\item What is the transmission mechanism of monetary policy?
\item Does forward guidance work?
\item What happens at the zero lower bound?
\item How effective is unconventional policy (QE)?
\end{itemize}

\textbf{Empirical strategies}
\begin{itemize}
\item High-frequency identification
\item Natural experiments
\item Central bank communication shocks
\end{itemize}

\section{What Causes Inflation Dynamics?}

\textbf{Central Question}

Why does inflation respond sluggishly to economic slack?

\textbf{Sub-questions}
\begin{itemize}
\item Is the Phillips Curve dead or nonlinear?
\item Role of expectations formation
\item Impact of supply shocks and globalization
\end{itemize}

\textbf{Frontier topics}
\begin{itemize}
\item Behavioral expectations
\item Inflation persistence
\item Sectoral price rigidity
\end{itemize}

\section{What Determines Long-Run Economic Growth?}

\textbf{Central Question}

Why do some countries grow faster than others over decades?

\textbf{Sub-questions}
\begin{itemize}
\item Role of technology and innovation
\item Human capital and education
\item Institutions and governance
\item Misallocation and productivity dispersion
\end{itemize}

\textbf{Methods}
\begin{itemize}
\item Growth accounting
\item Firm-level microdata
\item Endogenous growth models
\end{itemize}

\section{How Do Financial Frictions Shape Macroeconomic Outcomes?}

\textbf{Central Question}

How do credit markets amplify recessions and booms?

\textbf{Sub-questions}
\begin{itemize}
\item Role of bank lending constraints
\item Leverage cycles
\item Financial accelerator mechanisms
\item Macroprudential policy effectiveness
\end{itemize}

\textbf{Popular frameworks}
\begin{itemize}
\item Financial accelerator models
\item Intermediary asset pricing
\item Household balance sheet channels
\end{itemize}

\section{What Are the Macroeconomic Consequences of Inequality?}

\textbf{Central Question}

How does income and wealth inequality affect growth and stability?

\textbf{Sub-questions}
\begin{itemize}
\item Does inequality dampen consumption demand?
\item How do heterogeneous households respond to shocks?
\item What are distributional effects of monetary policy?
\end{itemize}

\textbf{Frontier approach}

Heterogeneous Agent New Keynesian (HANK) models.

\section{How Do Labor Markets Transmit Macro Shocks?}

\textbf{Central Question}

Why do unemployment and wages adjust slowly after shocks?

\textbf{Key issues}
\begin{itemize}
\item Search frictions and matching
\item Wage rigidity
\item Job polarization
\item Role of firm dynamics
\end{itemize}

\textbf{Data used}
\begin{itemize}
\item Matched employer--employee data
\item Administrative records
\end{itemize}

\section{How Do Expectations and Beliefs Affect Macro Outcomes?}

\textbf{Central Question}

Do expectations drive economic fluctuations?

\textbf{Sub-questions}
\begin{itemize}
\item Rational vs adaptive expectations
\item Information frictions
\item Belief heterogeneity
\end{itemize}

\textbf{Growing frontier}
\begin{itemize}
\item Survey expectations data
\item Bounded rationality models
\end{itemize}

\section{What Explains Rare Disasters and Deep Recessions?}

\textbf{Central Question}

Why do severe crises occur and what amplifies them?

\textbf{Topics}
\begin{itemize}
\item Financial crises
\item Pandemic macroeconomics
\item Sovereign debt crises
\item Systemic risk
\end{itemize}

\section{What Are the Macroeconomic Effects of Fiscal Policy?}

\textbf{Central Question}

How do government spending and taxation affect output and welfare?

\textbf{Key debates}
\begin{itemize}
\item Size of fiscal multipliers
\item State-dependent effects
\item Public debt sustainability
\item Fiscal policy at the zero lower bound
\end{itemize}

\section{Emerging Frontier Topics}

\subsection*{Climate Macroeconomics}
\begin{itemize}
\item Macroeconomic costs of climate change
\item Transition risks and green investment
\end{itemize}

\subsection*{Housing and the Macroeconomy}
\begin{itemize}
\item Housing cycles and financial stability
\item Mortgage markets and consumption
\end{itemize}

\subsection*{Supply Chains and Globalization}
\begin{itemize}
\item Resilience vs efficiency trade-offs
\end{itemize}

\subsection*{AI, Automation, and Productivity}
\begin{itemize}
\item Labor displacement vs productivity gains
\end{itemize}

\section{Common Features of Top-Journal Macro Papers}

Top macro papers typically:

\begin{itemize}
\item Ask quantitatively important questions
\item Combine theory with micro data
\item Use structural models or credible identification
\item Address policy relevance
\item Provide welfare analysis
\item Improve measurement or identification
\end{itemize}

\section{How the Frontier Has Evolved}

\textbf{1980s--1990s}
\begin{itemize}
\item RBC models
\item Technology shocks
\item Calibration
\end{itemize}

\textbf{2000s}
\begin{itemize}
\item New Keynesian DSGE
\item Monetary policy rules
\end{itemize}

\textbf{Post-2008 Crisis}
\begin{itemize}
\item Financial frictions
\item Macro-finance integration
\end{itemize}

\textbf{2015--Present}
\begin{itemize}
\item Heterogeneous agents
\item Expectations formation
\item Inequality and distribution
\item Climate and macro risk
\end{itemize}

\section{High-Impact Research Areas Given Your Expertise}

Given your background in global economic shifts, inequality, housing cycles, and labor markets, the following areas are highly publishable:

\begin{itemize}
\item Housing cycles and monetary tightening
\item Distributional effects of inflation shocks
\item Mortgage renewal risk and consumption smoothing
\item Construction labor dynamics and housing supply
\item Immigration shocks and regional housing inflation
\item AI adoption and labor market polarization
\end{itemize}

\end{document}