\documentclass[12pt]{article}

\usepackage[margin=1in]{geometry}
\usepackage{amsmath, amssymb}
\usepackage{booktabs}
\usepackage{setspace}
\usepackage{enumitem}

\setstretch{1.2}
\setlength{\parindent}{0pt}

\title{\textbf{In-Class Group Exercise}\\
Labour Markets and the Future of Work}
\author{ECON 204 -- Contemporary Economic Issues}
\date{}

\begin{document}

\maketitle

\section*{Instructions}

Work in groups of \textbf{2--3 students}. Answer the following questions carefully and be prepared to discuss your answers with the class. Show all calculations clearly where required.

Estimated time: \textbf{50--60 minutes}

\vspace{1em}

%====================================================
\section*{Part I: Structural Change in Employment}

The table below shows the employment shares of three major sectors in a hypothetical advanced economy.

\begin{center}
\begin{tabular}{lccc}
\toprule
Sector & 1990 & 2010 & 2025 \\
\midrule
Agriculture & 7\% & 4\% & 2\% \\
Manufacturing & 25\% & 16\% & 10\% \\
Services & 68\% & 80\% & 88\% \\
\bottomrule
\end{tabular}
\end{center}

\textbf{Questions}

\begin{enumerate}[label=\arabic*.]
\item Which sector experienced the largest decline in employment share between 1990 and 2025?
\vspace{1.5cm}

\item By how many percentage points did manufacturing employment fall over this period?
\vspace{1.5cm}

\item What broad structural change does this table illustrate about modern economies?
\vspace{2cm}

\item Give two economic explanations for the decline in manufacturing employment.
\vspace{2.5cm}

\item Why has service-sector employment become dominant in advanced economies?
\vspace{2.5cm}
\end{enumerate}

\newpage

%====================================================
\section*{Part II: Automation, Tasks, and Occupations}

Economists increasingly argue that technology replaces \textbf{tasks} rather than entire jobs.

Consider the following occupations:

\begin{center}
\begin{tabular}{lccc}
\toprule
Occupation & Mostly Routine? & Easily Automated? & Explain Briefly \\
\midrule
Accountant & & & \\
\vspace{0.5cm}\\
Taxi Driver & & & \\
\vspace{0.5cm}\\
Nurse & & & \\
\vspace{0.5cm}\\
Data Entry Clerk & & & \\
\vspace{0.5cm}\\
Software Developer & & & \\
\bottomrule
\end{tabular}
\end{center}

\textbf{Questions}

\begin{enumerate}[label=\arabic*.]
\setcounter{enumi}{5}
\item Which of the occupations above are most vulnerable to automation?
\vspace{2cm}

\item Why are routine tasks generally easier to automate?
\vspace{2cm}

\item Why are interpersonal, creative, and judgment-intensive tasks harder to automate?
\vspace{2cm}

\item Explain why automation may change a job without eliminating the occupation entirely.
\vspace{2.5cm}
\end{enumerate}

\newpage

%====================================================
\section*{Part III: Wage Inequality and Skill Demand}

Suppose the following wage data describe two categories of workers:

\begin{center}
\begin{tabular}{lcc}
\toprule
Worker Type & Average Wage in 2000 & Average Wage in 2025 \\
\midrule
High-skill workers & \$30/hour & \$52/hour \\
Low-skill workers & \$18/hour & \$22/hour \\
\bottomrule
\end{tabular}
\end{center}

\textbf{Questions}

\begin{enumerate}[label=\arabic*.]
\setcounter{enumi}{9}
\item Calculate the wage gap between high-skill and low-skill workers in 2000.
\vspace{1.5cm}

\item Calculate the wage gap between high-skill and low-skill workers in 2025.
\vspace{1.5cm}

\item Has wage inequality increased or decreased?
\vspace{1.5cm}

\item Calculate the percentage increase in wages for:
\begin{enumerate}[label=(\alph*)]
\item high-skill workers
\item low-skill workers
\end{enumerate}
\vspace{2.5cm}

\item Give two economic explanations for why the wage gap changed over time.
\vspace{2.5cm}
\end{enumerate}

\newpage

%====================================================
\section*{Part IV: A Simple Canadian Labour Market Data Exercise}

Suppose the following unemployment rates are reported for Canada:

\begin{center}
\begin{tabular}{lcc}
\toprule
Group & 2019 & 2025 \\
\midrule
Youth (15--24) & 11.0\% & 13.5\% \\
Prime-age workers (25--54) & 5.2\% & 6.0\% \\
Older workers (55+) & 4.8\% & 5.1\% \\
\bottomrule
\end{tabular}
\end{center}

\textbf{Questions}

\begin{enumerate}[label=\arabic*.]
\setcounter{enumi}{14}
\item Which group had the highest unemployment rate in both years?
\vspace{1.5cm}

\item By how many percentage points did youth unemployment increase between 2019 and 2025?
\vspace{1.5cm}

\item Why are younger workers often more vulnerable to labour market disruptions?
\vspace{2cm}

\item Explain why technological change may affect younger and older workers differently.
\vspace{2.5cm}
\end{enumerate}

\newpage

%====================================================
\section*{Part V: The Gig Economy and Non-Standard Work}

Many workers today earn income through digital platforms such as Uber, DoorDash, Instacart, or Upwork.

\textbf{Questions}

\begin{enumerate}[label=\arabic*.]
\setcounter{enumi}{18}
\item Give two advantages of gig work for workers.
\vspace{2cm}

\item Give two disadvantages of gig work for workers.
\vspace{2cm}

\item Should gig workers be treated as employees or independent contractors? Explain your reasoning.
\vspace{3cm}

\item What labour market policies could improve working conditions in the gig economy?
\vspace{3cm}
\end{enumerate}

\newpage

%====================================================
\section*{Part VI: AI and the Future of Work}

Read the following statement and answer the questions below.

\textit{``Artificial intelligence will not simply destroy jobs. Instead, it is likely to restructure many occupations by automating some tasks, complementing others, and increasing the value of workers who can adapt.''}

\textbf{Questions}

\begin{enumerate}[label=\arabic*.]
\setcounter{enumi}{22}
\item What does it mean to say that AI may \textbf{complement} labour rather than replace it?
\vspace{2cm}

\item Give one example of a job where AI could raise worker productivity.
\vspace{2cm}

\item Give one example of a job where AI could reduce labour demand.
\vspace{2cm}

\item What types of skills are likely to become more valuable in the future labour market?
\vspace{2.5cm}
\end{enumerate}

\newpage

%====================================================
\section*{Part VII: Policy Debate}

Choose \textbf{one} of the following policy responses:

\begin{itemize}
\item expanded retraining programs
\item stronger labour protections for gig workers
\item universal basic income
\item shorter work week
\item greater public investment in education and digital skills
\end{itemize}

Answer the following questions:

\begin{enumerate}[label=\arabic*.]
\setcounter{enumi}{26}
\item How would this policy affect workers?
\vspace{2cm}

\item How might this policy affect firms?
\vspace{2cm}

\item What unintended consequences might arise?
\vspace{2.5cm}

\item Would you recommend this policy? Why or why not?
\vspace{2.5cm}
\end{enumerate}

\vspace{1em}

\section*{Class Wrap-Up}

Each group should prepare a brief summary of:
\begin{itemize}
\item one key trend in labour markets,
\item one major risk from technological change,
\item one policy that could improve labour market outcomes in the future.
\end{itemize}

\end{document}