\documentclass[12pt]{article}
\usepackage[margin=1in]{geometry}
\usepackage{graphicx}
\usepackage{url}
\usepackage{hyperref}
\hypersetup{
    colorlinks=true,
    linkcolor=blue,
    filecolor=magenta,      
    urlcolor=cyan,
}
\usepackage{enumitem}
\setlist[itemize]{noitemsep, topsep=0pt} % Remove extra spacing in lists

\title{ECON 220: Global Economic Shifts \\ Term Project}
%\author{Department of Economics \\ University of Canada}
\date{Winter 2025}

\begin{document}

\maketitle

\section*{Objective}
The objective of this term project is to deepen students' understanding of global economic shifts by researching and analyzing a specific topic. Students will work in pairs to produce a comprehensive report and deliver a presentation, demonstrating their ability to critically evaluate economic trends, policies, and their implications.

\section*{Project Guidelines}

\subsection*{1. Group Formation}
\begin{itemize}
    \item Students will form groups of two.
    \item Each group will select one topic from the list provided below.
\end{itemize}

\subsection*{2. Report}
\begin{itemize}
    \item \textbf{Length:} 5--8 pages (excluding tables, figures, diagrams, and bibliography).
    \item \textbf{Format:} Double-spaced, 12-point font, 1-inch margins.
    \item \textbf{Structure:} The report should include the following sections:
    \begin{itemize}
        \item \textbf{Introduction:} Background information and significance of the topic.
        \item \textbf{Literature Review:} Summary of existing research and theories related to the topic.
        \item \textbf{Analysis:} In-depth analysis of the chosen topic, including current trends, data, and case studies.
        \item \textbf{Discussion:} Implications of the economic shift, potential future developments, and policy recommendations.
        \item \textbf{Conclusion:} Summary of key findings and their broader significance.
    \end{itemize}
    \item \textbf{Bibliography:} Properly cited sources using APA, MLA, or Chicago style.
\end{itemize}
\newpage
\subsection*{3. Presentation}
\begin{itemize}
    \item \textbf{Duration:} 20--30 minutes.
    \item \textbf{Format:} Each group will present their findings to the class using visual aids (e.g., PowerPoint).
    \item \textbf{Content:} The presentation should cover the main points of the report, including the introduction, analysis, discussion, and conclusion.
    \item \textbf{Q\&A:} Be prepared to answer questions from the instructor and classmates.
\end{itemize}

\subsection*{4. Suggested Topics}
\begin{itemize}
    \item The Impact of Digital Transformation on Global Labor Markets
    \item The Role of Cryptocurrencies in the Future of Global Finance
    \item Climate Change and the Transition to a Green Economy
    \item The Effects of Aging Populations on Economic Growth in Developed Countries
    \item The Rise of Protectionism and Its Impact on Global Trade
    \item The Economic Implications of Universal Basic Income (UBI)
    \item The Role of Emerging Markets in Shaping the Global Economy
    \item The Impact of Geopolitical Tensions on Global Supply Chains
    \item The Future of Work: Automation and Job Displacement
    \item The Economic Consequences of the COVID-19 Pandemic and Recovery Strategies
\end{itemize}

\subsection*{5. Deadlines}
\begin{itemize}
    \item \textbf{Topic Selection:} February 15, 2025

    \item \textbf{Draft:} March 15,  2025
    \item \textbf{Final Report Submission:} March 31, 2025
    \item \textbf{Presentations:} March 17 to April 17,  2025 (Exact dates to be assigned)
\end{itemize}

\subsection*{6. Evaluation Criteria}
\begin{itemize}
    \item \textbf{Report (60\%):}
    \begin{itemize}
        \item Depth of research and analysis (20\%)
        \item Clarity and organization of content (15\%)
        \item Use of data and evidence (15\%)
        \item Quality of writing and citation (10\%)
    \end{itemize}
    \item \textbf{Presentation (40\%):}
    \begin{itemize}
        \item Clarity and coherence of presentation (15\%)
        \item Engagement and delivery (10\%)
        \item Visual aids and professionalism (10\%)
        \item Ability to answer questions (5\%)
    \end{itemize}
\end{itemize}
\newpage
\subsection*{7. Resources}
\begin{itemize}
    \item Access to academic journals, books, and reputable online sources.
    \item Guidance from the instructor during office hours or scheduled meetings.

\end{itemize}

\subsection*{8. Submission}
\begin{itemize}
    \item Reports should be submitted electronically via Moodle.
    \item Presentations will be scheduled during the final weeks of classes.
\end{itemize}

\section*{Additional Notes}
\begin{itemize}
    \item Collaboration is key. Ensure both group members contribute equally to the research, writing, and presentation.
    \item Start early to allow ample time for research, writing, and revisions.
 
\end{itemize}


\section*{Tips for Success}
\begin{itemize}
    \item Start early and divide tasks equally between group members.
    \item Use reliable sources such as Statistics Canada, academic journals, and government reports.
    \item Practice your presentation to ensure it fits within the time limit and flows smoothly.

\end{itemize}

\section*{Additional Resources}
\begin{itemize}
    \item Statistics Canada: \url{https://www.statcan.gc.ca/}
    \item Bank of Canada: \url{https://www.bankofcanada.ca/}
    \item Government of Canada Publications: \url{https://www.canada.ca/en/services/finance/publications.html}
\end{itemize}



\vspace{2cm}
\centering
\textbf{Good luck!} I look forward to seeing your insights on these important economic issues. If you have any questions, feel free to reach out.
\end{document}