\documentclass[11pt]{article}
\usepackage[margin=1in]{geometry}
\usepackage{amsmath, amssymb, amsthm, bm, mathtools}
\usepackage{booktabs}
\usepackage{array}
\usepackage{enumitem}

\title{ECON 101: Macroeconomics \\ Quiz 2}
\author{Instructor: Muhebullah Karimzada}
\date{October 01,  2025}

\begin{document}
\maketitle



\section*{Instructions}
Answer all questions on the answer sheet.  Show all work for numerical problems. Multiple-choice questions have one correct answer.


\section*{Part A – Multiple Choice (1 mark each)}

\textbf{1.} Suppose beef and chicken are substitutes. If the price of beef rises, holding all else constant:  
\begin{enumerate}[label=\alph*)]
    \item Demand for chicken decreases.  
    \item Supply of chicken decreases.  
    \item Demand for chicken increases.  
    \item Quantity demanded of chicken decreases.  
\end{enumerate}

\bigskip

\textbf{2.} Which of the following transactions is \textbf{included} in Canada’s GDP?  
\begin{enumerate}[label=\alph*)]
    \item A Canadian student buys used textbooks from another student.  
    \item A Canadian firm sells lumber to a U.S. company.  
    \item A Canadian tourist purchases a meal in Paris.  
    \item A Canadian household does its own home repairs.  
\end{enumerate}

\bigskip

\textbf{3.} Real GDP differs from nominal GDP because real GDP:  
\begin{enumerate}[label=\alph*)]
    \item Includes only goods, not services.  
    \item Is adjusted for population size.  
    \item Values output at constant prices.  
    \item Excludes investment spending.  
\end{enumerate}

\bigskip
\newpage
\textbf{4.} If the GDP deflator increases from 120 to 132, the inflation rate is approximately:  
\begin{enumerate}[label=\alph*)]
    \item 10\%  
    \item 12\%  
    \item 9\%  
    \item 8\%  
\end{enumerate}

\bigskip

\textbf{5.} Which of the following would \textbf{not} increase GDP?  
\begin{enumerate}[label=\alph*)]
    \item Higher consumption of new cars.  
    \item More exports of manufactured goods.  
    \item An increase in unpaid household work.  
    \item Greater investment in factories.  
\end{enumerate}

\bigskip

\section*{Part B – Numerical Problems (15 marks total)}

\textbf{1. Market Equilibrium (4 marks)}  
The market demand and supply for pizzas in Toronto are given by:  
\[
Q_D = 200 - 5P, \quad Q_S = 20 + 3P,
\]  
where \(Q\) is the number of pizzas and \(P\) is the price in dollars.  

\begin{enumerate}[label=\alph*)]
    \item Find the equilibrium price and quantity.  
    \item If the government imposes a price floor at \$25, calculate the resulting surplus or shortage.  
\end{enumerate}

\bigskip

\textbf{2. Expenditure Approach to GDP (4 marks)}  
Suppose the following Canadian macroeconomic data (in billions):  

\begin{itemize}
    \item Consumption: \$1,200  
    \item Investment: \$500  
    \item Government spending: \$600  
    \item Exports: \$450  
    \item Imports: \$550  
\end{itemize}

\begin{enumerate}[label=\alph*)]
    \item Calculate nominal GDP using the expenditure approach.  
    \item If population = 37 million, compute GDP per capita.  
    \item If potential GDP = \$3,000 billion, calculate the output gap as a percentage.  
\end{enumerate}

\bigskip

\textbf{3. Real vs. Nominal GDP (7 marks)}  
An economy produces two goods: milk and computers.  

\begin{center}
\begin{tabular}{c c c c c}
\toprule
Year & Quantity Milk & Price Milk & Quantity Computers & Price Computers \\
\midrule
2023 & 1,000 & \$2 & 50 & \$500 \\
2024 & 1,200 & \$3 & 55 & \$550 \\
\bottomrule
\end{tabular}
\end{center}

\begin{enumerate}[label=\alph*)]
    \item Compute nominal GDP for 2023 and 2024.  
    \item Compute real GDP for 2024 using 2023 as the base year.  
    \item Calculate the GDP deflator for 2024.  
    \item What is the inflation rate between 2023 and 2024?  
\end{enumerate}

\end{document}
