\documentclass[12pt]{article}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{amsmath, amssymb, geometry, booktabs}
\usepackage{enumitem}
\geometry{margin=1in}
\setlength{\parindent}{0pt}
\renewcommand{\arraystretch}{1.2}


\title{Tutorial 2 --The Instruments of Trade Policy}
\author{ECON 308 -- International Economic Relations}
\date{Fall 2025}

\begin{document}
\maketitle



\subsection*{Problem 1 -- Tariff and Welfare for a Small Country}

Home imports wheat at a world price \(P_W = \$200\) per ton.  
Domestic supply and demand functions are:
\[
Q_S = P - 100, \qquad Q_D = 700 - P,
\]
where \(Q\) is measured in tons and \(P\) in dollars.

\begin{enumerate}[label=(\alph*)]
\item Compute equilibrium imports under free trade.  
\item The government imposes a \textbf{specific tariff} of \$20 per ton.  
Determine the new domestic price, domestic quantity supplied, quantity demanded, and imports.
\item Calculate tariff revenue.  
\item Compute the deadweight losses from both production and consumption distortions.
\item Determine the net welfare loss.
\end{enumerate}

\vspace{1.2em}

\subsection*{Problem 2 -- Large Country and the Terms of Trade Effect}

Home faces a foreign export supply curve:
\[
P^* = 100 + 0.5Q^*,
\]
where \(P^*\) is the foreign export price (in dollars per ton).  
Home's import demand is:
\[
Q_M^D = 400 - 2P,
\]
where \(P\) is the domestic import price.

\begin{enumerate}[label=(\alph*)]
\item Find the free-trade equilibrium price and quantity of imports.  
\item Home imposes a tariff of \$20 per ton.  
Compute the new world price, the new domestic price, and the volume of imports.
\item Calculate tariff revenue and the resulting terms-of-trade gain for Home.  
\item Determine whether welfare rises or falls by comparing the terms-of-trade gain to the deadweight losses.
\item Briefly explain why retaliatory policies from trading partners may prevent welfare gains in reality.
\end{enumerate}


\subsection*{Problem 3 – Effective Rate of Protection}

A car sells internationally for \$20{,}000.  
Its imported engine costs \$8{,}000 and is subject to a 25\% tariff.  
Finished cars face a 10\% tariff.

\begin{enumerate}[label=(\arabic*)]
\item Compute the \textbf{nominal protection rate} on the finished car.
\item Compute the \textbf{effective rate of protection (ERP)}:
\[
\text{ERP} = \frac{V_A' - V_A}{V_A},
\]
where domestic value added is \(V_A\) without tariffs and \(V_A'\) with tariffs.
\item Interpret whether the industry’s effective protection is greater or less than the nominal tariff.
\end{enumerate}


\end{document}
