\documentclass{beamer}
\usepackage{graphicx}
\usepackage{booktabs}

% Theme and title setup
\usetheme{Frankfurt}
\title{Introduction to Contemporary Economic Issues in Canada and Globally}
\subtitle{Contemporary Economic Issues: ECON 204}
\author{Muhebullah Karimzada}
\date{Winter 2025}


\begin{document}

% Title slide
\begin{frame}
  \titlepage
\end{frame}

% Slide: Course Overview
\begin{frame}{Course Overview}
  \begin{itemize}
    \item This course examines key economic challenges and opportunities in Canada and the global economy.
    \item Topics include globalization, inequality, technological change, climate economics, and post-pandemic recovery.
    \item The course integrates theory, empirical analysis, and policy discussions.
  \end{itemize}
\end{frame}

% Slide: Importance of Contemporary Economics
\begin{frame}{Why Study Contemporary Economics?}
  \begin{itemize}
    \item Understand the forces shaping modern economies.
    \item Analyze economic policies affecting society, businesses, and individuals.
    \item Address pressing global and national challenges, such as climate change and inequality.
    \item Develop critical thinking and data analysis skills applicable in policymaking and industry.
  \end{itemize}
\end{frame}

% Slide: Key Themes
\begin{frame}{Key Themes of the Course}
  \begin{itemize}
    \item \textbf{Globalization and Trade:} Impact on Canada and global economies.
    \item \textbf{Income Inequality:} Trends, causes, and policy responses.
    \item \textbf{Climate Change Economics:} Balancing growth with sustainability.
    \item \textbf{Technological Disruption:} Automation, AI, and labor markets.
    \item \textbf{Post-Pandemic Economic Recovery:} Fiscal and monetary challenges.
  \end{itemize}
\end{frame}

% Slide: Globalization and Trade
\begin{frame}{Globalization and Trade}
  \begin{itemize}
    \item \textbf{Definition:} The increasing integration of economies through trade, investment, and technology.
    \item \textbf{Impacts on Canada:}
      \begin{itemize}
        \item Opportunities for export-driven growth (e.g., natural resources, technology).
        \item Challenges: Competition from low-cost producers and trade disputes.
      \end{itemize}
    \item \textbf{Global Perspective:}
      \begin{itemize}
        \item Rising protectionism (e.g., tariffs, trade wars).
        \item Role of international institutions like WTO, IMF.
      \end{itemize}
  \end{itemize}
\end{frame}

% Slide: Income Inequality
\begin{frame}{Income Inequality}
  \begin{itemize}
    \item \textbf{Trends:} Growing disparities in income and wealth globally and in Canada.
    \item \textbf{Causes:}
      \begin{itemize}
        \item Technological changes favoring skilled labor.
        \item Globalization and outsourcing.
        \item Declining unionization and policy shifts.
      \end{itemize}
    \item \textbf{Policy Responses:}
      \begin{itemize}
        \item Progressive taxation and social welfare programs.
        \item Education and workforce development.
      \end{itemize}
  \end{itemize}
\end{frame}

% Slide: Climate Change Economics
\begin{frame}{Climate Change Economics}
  \begin{itemize}
    \item \textbf{Global Challenge:} Reducing emissions while supporting economic growth.
    \item \textbf{Canada’s Role:}
      \begin{itemize}
        \item Transition to renewable energy.
        \item Policies: Carbon pricing, green technology investments.
      \end{itemize}
    \item \textbf{Economic Opportunities:}
      \begin{itemize}
        \item Job creation in renewable sectors.
        \item Leading global innovation in sustainable technologies.
      \end{itemize}
  \end{itemize}
\end{frame}

% Slide: Technological Disruption
\begin{frame}{Technological Disruption}
  \begin{itemize}
    \item \textbf{Automation and AI:}
      \begin{itemize}
        \item Potential for increased productivity.
        \item Risks of job displacement in traditional industries.
      \end{itemize}
    \item \textbf{Implications for Policy:}
      \begin{itemize}
        \item Need for reskilling and education programs.
        \item Balancing innovation with ethical considerations.
      \end{itemize}
    \item \textbf{Global Competition:} Role of tech giants and innovation hubs.
  \end{itemize}
\end{frame}

% Slide: Post-Pandemic Recovery
\begin{frame}{Post-Pandemic Economic Recovery}
  \begin{itemize}
    \item \textbf{Challenges:}
      \begin{itemize}
        \item High inflation and debt levels.
        \item Supply chain disruptions.
      \end{itemize}
    \item \textbf{Opportunities:}
      \begin{itemize}
        \item Investment in healthcare and technology.
        \item Rethinking global supply chains.
      \end{itemize}
    \item \textbf{Policy Responses:}
      \begin{itemize}
        \item Monetary policy: Interest rates, quantitative easing.
        \item Fiscal policy: Stimulus packages and infrastructure investments.
      \end{itemize}
  \end{itemize}
\end{frame}

% Slide: Discussion Questions
\begin{frame}{Discussion Questions}
  \begin{itemize}
    \item How has globalization impacted Canada’s economy? Provide examples.
    \item What are the primary drivers of income inequality in Canada and globally?
    \item How can Canada balance economic growth with environmental sustainability?
  \end{itemize}
\end{frame}

% Slide: Critical Thinking Questions
\begin{frame}{Critical Thinking Questions}
  \begin{itemize}
    \item Can Canada remain competitive in a rapidly changing global economy? Why or why not?
    \item To what extent should Canada rely on natural resources for economic growth?
    \item What role does technology play in shaping future economic policies?
  \end{itemize}
\end{frame}

% Slide: Conclusion and Next Steps
%\begin{frame}{Conclusion and Next Steps}
 % \begin{itemize}
 %   \item Contemporary economics addresses pressing issues in Canada and globally.
 %   \item Engage with the material, participate in discussions, and apply critical thinking.
   % \item Next lecture: \textbf{Economic Models and Tools for Analyzing Issues.}
 % \end{itemize}
%\end{frame}

\end{document}

